---
title: "ESM 244 Lab 1"
author: "Katie Munster"
date: "1/11/2022"
output: html_document
---

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(here)
library(sf) # install.packages('sf')
library(tmap)
### install.packages('tmap')
### update.packages(ask = FALSE)
```

cmd-shift-enter shortcut for running the current code chunk

### Read in the data

cmd-option-i shortcut for creating a code chunk

```{r}
sf_trees <- read_csv(here('data', 'sf_trees', 'sf_trees.csv'),
                     show_col_types = FALSE)
```

names(sf_trees) to get the names of the columns
summary(sf_trees) to get an idea about the data in each of those columns

### Part 1: wrangling and ggplot review

**Example 1:** Find counts of observation by `legal_status` & wrangle a bit.

```{r}
### method 1: group_by() %>% summarize()
sf_trees %>% 
  group_by(legal_status) %>% 
  summarize(tree_count = n())

### method 2: different way plus a few new functions
### store stuff in an object this time
### highlight code and press command+enter
top_5_status <- sf_trees %>% 
  count(legal_status) %>% 
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% 
  slice_max(tree_count, n = 5) %>% 
  arrange(desc(tree_count))
### drop_na() would get rid of a row with NA in any column. we just want NA in legal_status column not for example na in a date column
### rename column from n to tree_count
### relocates that column to the front of the dataframe
### look at all values of tree_count and slice out the highest ones. in this case the top 5
### arrange(tree_count) does lowest to highest
### to get highest to lowest: arrange(-tree_count) OR arrange(desc(tree_count))
```

Make a graph of the top 5 from above:

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
  geom_col(fill = 'darkgreen') +
  labs(x = 'Legal status', y = 'Tree count') +
  coord_flip() +
  theme_minimal()
### to change the order of the columns: fct_reorder(legal_status, tree_count) to order the trees by tree_count. if wanted largest to smallest, then do -tree_count
```

**Example 2:** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

shift-cmd-c to comment/uncomment quickly

```{r}
# sf_trees$legal_status %>% unique()
# unique(sf_trees$caretaker)
permitted_data_df <- sf_trees %>% 
  filter(legal_status == 'Permitted Site', caretaker == 'MTA')
### can also use & instead of ,
### use | as or
### filter(legal_status %in% c('Permitted Site', 'Private') & caretaker == 'MTA')
```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal_status`, `date`, `latitude`, `longitude` and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>% 
  filter(str_detect(species, 'Blackwood Acacia')) %>% 
  select(legal_status, date, lat = latitude, lon = longitude)
### rename column to lat

### Make a little graph of locations
ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) +
  geom_point(color = 'darkgreen')
### doesn't know this is spatial data yet
```

**Example 4:** use `tidyr:separate()`

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, into = c('spp_scientific', 'spp_common'), sep = ' :: ')
```

**Example 5:** use `tidyr::unite()`

```{r}
ex_5 <- sf_trees %>% 
  unite('id_status', tree_id, legal_status, sep = '_COOL_')
```




